---
title: "R Training 4"
author: "Rachel Kenny"
date: "5/7/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages & data
```{r load}

# Load packages
library(readxl)
library(readr)
library(tidyverse)
library(janitor)

# Load agua chinon veg data
ac_data_raw <- read_excel("OCWR_AC_2019_Data.xlsx")
ac_data <- clean_names(ac_data_raw)
ac_data$polygon_id[which(ac_data$polygon_id=="5M Buffer")] <- "5M BUFFER"

# Load weir oak restoration data
oak_data_raw <- read_csv("Weir_Oak_Restoration_Data_winter19_2.csv")
oak_data <- clean_names(oak_data_raw) 

# Load original oak data
oak_data_old <- read_csv("Weir_Oak_Restoration_Data_winter19.csv") %>% clean_names(oak_data_old) 

# Load arthropod data
arth_data <- read_csv("IRC_restoration_arthropod_data.csv") %>% 
  clean_names()

View(arth_data)

```

## Subsetting data
+ my_df[1:3] (no comma) will subset my_df, returning the first three columns as a data frame.

+ my_df[1:3, ] (with comma, numbers to left of the comma) will subset my_df and return the first three rows as a data frame

+ my_df[, 1:3] (with comma, numbers to right of the comma) will subset my_df and return the first three columns as a data frame, the same as my_df[1:3]

## Common operators
+ != indicates Not equal to
+ !X indicates Not X
+ X|Y indicates X or Y
+ X&Y indicates X and Y

## Practice subsetting data
```{r subset}

# To return rows 1-3
ac_data[1:3]

# To return columns 1-3
ac_data[1:3,]

# To return rows 1-3 (same as first option)
ac_data[,1:3]

# Single brackets extract a list
ac_data[5]

# Double brackets extract the items in a list
ac_data[[5]]

```

## If else statements
If else statements work similarly in r as they do in excel. There is a logical test given, then a value to be given if the logical statement is true and a value given if false.

df$new_column <- ifelse(test, value if true, value if false)

If else statements can also be nested within one another. The second if else statement is placed where the "value if false" goes.

df$new_column <- ifelse(test, value if true, (ifelse(test2, value2 if true, value2 if false))


## If else statements
```{r ifelse}

# Create a duplicate of our dataframes
ac_data2 <- ac_data
oak_data2 <- oak_data

# Ifelse statement
ac_data2$new_column <- ifelse(ac_data2$native_non_native == "Non-Native" | ac_data2$native_non_native == "No_Native", 0, 1)

oak_data2$new_column <- ifelse(oak_data2$height_cm >= 5, "Tall", "Short")

View(ac_data2)
View(oak_data2)

# Nested ifelse statement
ac_data2$new_column2 <- ifelse(ac_data2$native_non_native == "Native", 1, (ifelse(ac_data2$native_non_native == "Non-Native", 2, 3)))

oak_data2$new_column2 <- ifelse(oak_data2$height_cm >= 10, "Tall", (ifelse(oak_data2$height_cm > 0 & oak_data2$height_cm < 10, "Short", "No Germination")))

View(ac_data2)
View(oak_data2)

```


## Basic statistics
```{r stats}

# Mean
mean(oak_data$height_cm)

# Median
median(oak_data$height_cm)

# Standard deviation
sd(oak_data$height_cm)

# Highest number
max(oak_data$height_cm)

# Lowest number
min(oak_data$height_cm)

# Summarize data
summary(oak_data)

```


## Basic statistics - troubleshooting
```{r troubleshoot}

# Mean
mean(oak_data_old$height_cm)

# If there are NA values, then introduce na.rm = TRUE to exclude those values from analysis
mean(oak_data_old$height_cm, na.rm = TRUE)

# Standard deviation
sd(oak_data_old$height_cm, na.rm = TRUE)

# Highest number
max(oak_data_old$height_cm, na.rm = TRUE)

# Lowest number
min(oak_data_old$height_cm, na.rm = TRUE)

# Identify problem rows
which(is.na(oak_data_old$height_cm)) #819
which(grepl(834, oak_data_old$height_cm)) #933

# Edit data
# Method 1
oak_data_new <- oak_data_old[-c(819, 933), ] 
View(oak_data_new)

# Method 2
oak_data_new <- oak_data_old[ !(oak_data_old$short_id %in% c("U_C_33_1", "U_C_04_3")), ]
View(oak_data_new)

# It is always safer to delete data according to some sort of identifying value than to delete by row number. Row numbers can change as data is filtered and altered and it is therefor not the safest

```


## Tidyr - Spread and gather
A table is considered tidy data if each variables is in it's own column, each row represents one observation, and each cell has a single value. It is sometimes necessary to rearrange the data in order to suit the needs of the analysis or data visualization.

Data is rearranged according to the key and the value. The key is an attribute and explains what the data describes. The value is the value of the observation described by the key.

+ Spread() is used to seperate data from one column into many
+ Gather() is used to take two columns, the key column and value column, and seperate them into many columns based on the key column attributes

Spread(data, key, value)
Gather(data, key, value)

Check out these links for more information:
https://garrettgman.github.io/tidying/


## Spread and gather examples
```{r spread_gather}

# Gather - we are taking all of the proportional biomass values and placing them in one column
arth_data_2 <- arth_data %>% 
  select(transect, year_restoration, detritivores, herb_chew, herb_suck, parasitoid, pollinator, predator, scavenger) %>% 
  gather("functional_group", "proportional_biomass", 3:9)

View(arth_data_2)

# Spread - we are taking all of the proportional biomass and seperating them back out by their type
arth_data_3 <- arth_data_2 %>% 
  spread("functional_group", "proportional_biomass")

View(arth_data_3)

# By treating restoration year as a factor, we can have the graph aligned so that years 0, 1, 3, and 5 are equidistant
arth_data_2$year_restoration <- as.factor(arth_data_2$year_restoration)

# Create a ggplot
plot_arth <- ggplot(arth_data_2) + 
  geom_bar(aes(x = year_restoration, y = proportional_biomass, fill = functional_group), position = "fill", stat = "summary", fun.y = "mean") +
  theme_classic() +
  ggtitle("Changes in proportional biomass of arthropods over time") +
  xlab("Year of restoration") +
  ylab("Proportional biomass") +
  scale_y_continuous(expand=c(0,0), breaks = c(.2,.4,.6,.8,1))+
  scale_x_discrete(expand=c(0.2,0))+
  scale_fill_brewer("Functional Group",palette="Set2", labels=c("Detritivore", "Herbivore (chewing)", "Herbivore (sucking)", "Parasitoid", "Pollinator", "Predator", "Scavenger")) +
  theme(plot.title = element_text(face="bold", size = 11),
        axis.title = element_text(size=8),
        legend.title = element_text(size=8, face="bold"),
        legend.text = element_text(size=8))

plot_arth

```


